The Spatial and Temporal Domains of Modern Ecology

### 2017-06-22 paper/main.tex
Near final submission (JJC, PE, TT, JC edits)

%\begin{sciabstract}
\section*{Abstract}
%Ecologists have long known that the observational scales play a critical role in shaping our understanding of ecosystems, and that observations spanning multiple scales of space and time are needed to effectively characterize ecological phenomena. 
\textbf{To properly understand ecological phenomena, it is necessary to observe them across a range of spatial and temporal scales. Ecologists first raised this point in the 1980s, since when the ability to collect multi-scale observations has grown rapidly. To assess modern ecology's progress in addressing scale, we analyzed the resolution, extent, interval, and duration of observations in 348 studies published between 2004-2014. We found that the scale domains of observations remain fairly narrow, and are collected primarily with conventional field techniques. Compared to prior meta-analyses from the 1980s and early 2000s, observational durations and resolutions remain largely unchanged, but intervals have become finer and extents larger. Despite such gains, a large gulf exists between the scales at which phenomena are actually observed, and the scales those observations ostensibly represent, raising questions about observational representativeness. Adding to this concern, observational scales were not clearly reported in most studies, suggesting that scale often is a minor consideration. Journals can help mitigate this problem by adopting scale reporting standards, which can further spur ecologists to adopt rapidly improving observational technologies, and thereby close key gaps in current observational domains.}

@@

Our analysis focused on two dimensions of spatial scale, resolution (grain) and extent, and two of temporal scale, interval and duration (Table 1). Resolution is the area of an individual spatial replicate within which a complete measurement (as opposed to a sub-sample) of the feature of interest was made. Extent is the area enclosed by the outer-most spatial replicates, or, if the system or habitat being sampled was distinct from its surrounding matrix (e.g. forest patches in grassland habitats), the summed area of sampled patches (see SI for full definition). Interval refers to the average time elapsed between individual temporal replicates. Duration measures the time elapsed between the first and last temporal replicates, or, in the case of temporally unreplicated observations, the estimated time spent collecting the observation (SI). We also assessed observational scales within two additional dimensions, \emph{actual} extent (the integrated area of spatial replicates) and \emph{actual} duration (the summed observational time of temporal replicates). We evaluated these additional dimensions to gain insight into how much the actual scales of observation (i.e. how much space and time is covered by the measurement) differ from those they are intended to represent.

@@

\noindent \textbf{Results}\\
\noindent \textbf{\emph{Domains}}\\
In terms of resolution, the majority (67\%) of observations were collected in plots of $<$1 m$^2$ resolution, 24\% were collected within plots of 1 m$^2$ up to 1 ha, and the remaining 9\% in plots of $\geq$1 ha (Fig. 1A). The extent of 19\% of observations was $<$10 ha, 23\% covered 10-1,000 ha, 11\% 1-10,000 ha, 19\% 10,000-100,000 ha, 27\% 100,000-1,000,000 ha, and 16\% $>$1,000,000 ha (Fig. 1B).

@@

Comparing the two spatial dimensions against one another (for all observations) shows a primary concentration of observations with 10 cm$^2$ to 100 m$^2$ resolution that have extents ranging between slightly over 1,000 to nearly 1,000,000 ha (Fig. 2C). The second-most prominent concentration consists of higher resolution (1 cm$^2$-1 m$^2$), smaller extent (10-1,000 ha) observations, beneath which lies a third and fainter concentration of 1-1,000 cm$^2$ resolution, 1000 m$^2$ to $<$10 ha.  These three concentrations suggest a tendency for observational extent to increase with resolution, which is a relationship that becomes more pronounced in the (less densely observed) portion of the domain where resolutions $\geq$100 m$^2$.

The majority of temporally replicated observations have daily to decadal intervals and durations of $\geq$ 1 month to 1 decade. The orientation of this concentration shows that interval increases with duration; observations lasting one month to one year tend to have daily to monthly intervals, while those lasting one year to one decade tend to have yearly to decadal intervals. The low densities of observations having sub-daily intervals shows that relatively few high frequency, long duration ecological measurements are undertaken.

To provide further insight into observational domains, we also evaluated the degree to which the scales \emph{represented} by observations differed from those that were actually observed. To make this assessment, we first log$_{10}$ transformed and then subtracted the values of i) actual extent from extent and ii) actual duration from duration, in order to calculate the magnitude of difference (or decade) between each pair of dimensions for each observation. We then compared how these magnitudes varied in relation to the scale of the actual dimension (Fig. 3). This comparison showed that a majority (81\%) of the assessed observations had actual extents of $\leq$1 ha, which on average were 4 to nearly 8 orders of magnitude smaller than the quantified extent (Fig 3.A).  Actual extent converged with extent $\geq$1 100,000 ha, but this applies to just 5\% of observations. The actual duration of 64\% of observations was $<$ 1 day, which on average was 3 to nearly 9 orders of magnitude shorter than the time span covered by temporal replicates (Fig. 3B). The two duration measures were approximately equal for the 16\% of observations exceeding one month of actual duration.

@@

\noindent \textbf{\emph{Potential biases and uncertainties in quantifying scales}}\\
There were several potential methodological aspects that could have influenced our assessment of ecology's spatial and temporal domains. The first stems from our finding that many studies did not precisely report observational scales, which meant that we had to estimate--rather than simply record--these values for most observations (specifically, in 63, 60, and 69\% of cases for resolution, extent, and actual extent, and 36\%, 64\%, and 83\% of cases for interval, duration, and effective duration, respectively). The inevitable estimation errors may have biased our overall findings. However, we attempted to quantify this error by assessing inter-rater disagreement and incorporating this uncertainty into our resampling methodology. The resulting confidence intervals (Fig. 1) suggest that it was unlikely that estimation errors unduly influenced our findings.

@@

%Adding to our motivation to record actual observational scales was the fact that 
\noindent \textbf{Discussion}\\
\noindent \textbf{\emph{Insights into the scale domains of modern ecology}}\\
Our results suggest that the scale domains of modern ecological observations are fairly narrow, and are collected primarily with conventional field techniques. Spatially, the majority of observations have grains of $\leq$1m$^2$ and extents of $\leq$10,000 ha (Fig 1A;B). In the temporal domains, most observations are either un-replicated snapshots, or of low frequency ($\geq$1 month interval; Fig. 1C) and relatively short duration ($\leq$1 year; Fig. 1D; 2D). Contrasting observational dimensions reveals that larger extents are associated with larger plot sizes (Fig. 2C), while longer durations are associated with longer intervals (Fig. 2D). The latter association presumably reflects a cost-imposed tradeoff between sampling frequency and temporal duration that is characteristic of traditional field-based observation. The same tradeoff is also responsible for the inverse relationship between resolution and interval that dominates that domain space (Fig. 2A). As a result of these tradeoffs, there are notable ``holes'' in the domains defined by high frequency (daily to sub-daily intervals) observations having 1) high to moderate resolutions ($\geq$1 m$^2$ up to 100 ha; Fig. 2A) and 2) decadal or longer durations (Fig. 2B).

@@

Have these domains changed since the seminal papers on scale first began to appear in the late 1980s \cite{wiens_spatial_1989, levin_problem_1992, tilman_ecological_1989}? A comprehensive answer to this question is not possible, absent a similar study targeting earlier literature, but an analysis of results presented in three prior studies provides partial insight. The first and most comprehensive dataset consists of duration values extracted by Tilman \cite{tilman_ecological_1989} from 623 studies published between 1977-1987 in the journal \emph{Ecology}. The average duration of the most comparable subset of those values (n=419; see SI) was 3.6 years, compared to 3.3 years for observations in our sample (or 5.1 if temporally un-replicated observations are excluded). The second dataset is found in Kareiva and Anderson \cite{kareiva_spatial_1988}, who present the resolutions of 97 community ecology experiments published in \emph{Ecology} between 1980-1986. The average of those (12,657 m$^2$) was substantially smaller than the mean of our sample (1,510,247 m$^2$), but comparing the 80th percentile value (197 m$^2$) of Kareiva and Anderson's \cite{kareiva_spatial_1988} to ours (197 m$^2$) shows that the majority of contemporary observations are finer-grained than most 1980s-era experiments. The third dataset is provided by Porter et al \cite{porter_wireless_2005}, who compared the extent and interval of 25 studies published in 2003 and 2004 (also in \emph{Ecology}). The mean interval was 178 days, compared to 684 in our sample, but the 80th percentile value in our study was 169 compared to 329 in the earlier study. Extent in our sample was substantially larger according to multiple summary statistics, including the mean (115,098,283 ha in our study versus 368,403 ha), median (5,120 ha versus 9 ha), and 95th percentile (45,104,594 ha versus 136,000 ha).

Although limited due to methodological differences (e.g. a focus on experiments versus un-manipulated systems; SI), these comparisons suggest that the duration and, less clearly, resolution of ecological observations have changed little in the past 30 years, but observational frequency and particularly extent have increased. The growth in observational extent is also reflected by the positive year-on-year trend within our own dataset, which corresponds to the increasing use of remote sensing (SI).

@@

It is therefore important to examine whether the phenomena being observed are adequately captured by the design of replicates. Our methods suggest one possible procedure for assessing the representativeness of replicates: 1) measure auto-correlation (spatial or serial) in the replicates, 2) add the autocorrelation length to the replicate area/duration, 3) calculate an autocorrelation-adjusted actual extent/duration, and 4) plot where it falls between actual extent/duration and extent/duration. If the adjusted actual value is substantially closer to the representative value, then this may provide greater confidence in the adequacy of the sampling scheme. If not, then alternative sampling methods may be used to close this gap. For example, remote sensing provides wall-to-wall spatial coverage of a study area, erasing the difference between actual extent and extent. Similarly, the interval of high-resolution (a desirable image property because it allows individual ecological features to be more effectively discerned  \cite{dark_modifiable_2007, hay_comparison_2003}) imaging is now approaching daily to sub-daily scales \cite{drusch_sentinel-2:_2012, hand_startup_2015}, allowing improved representation of temporal dynamics. For phenomena that can't be measured from space, either because they are not visible or because they require continuous observation, new approaches for \emph{in situ} or near-surface observations (e.g. low-cost wireless sensors \cite{wolf_gsm-based_2012, collins_new_2006, porter_wireless_2005}, citizen observers \cite{dickinson_current_2012}, and autonomous vehicles \cite{anderson_lightweight_2013}) can be used to increase the spatial and temporal coverage of observations.

>>>>

%\begin{sciabstract}
\section*{Abstract}
%Ecologists have long known that the observational scales play a critical role in shaping our understanding of ecosystems, and that observations spanning multiple scales of space and time are needed to effectively characterize ecological phenomena. 
\textbf{To properly understand ecological phenomena, it is necessary to observe them across a range of spatial and temporal scales. Ecologists first raised this point in the 1980s, and since then the ability to collect multi-scale observations has grown rapidly. To assess modern ecology's progress in addressing scale, we analyzed the resolution, extent, interval, and duration of observations in 348 studies published between 2004-2014. We found that the scale domains of observations are fairly narrow, and are collected primarily with conventional field techniques. In the spatial domain, most observations have resolutions of $\leq$1 m2 and extents of $\leq$10,000 ha. In the temporal domain, most observations were either unreplicated or of low frequency ($\geq$1 month interval), and were made over relatively short durations ($\leq$1 year). Compared to prior meta-analyses from the 1980s and early 2000s, observational durations and resolutions remain largely unchanged, but intervals have become finer and extents larger. Despite such gains, a large gulf exists between the scales at which phenomena are actually observed, and the scales those observations ostensibly represent, raising concerns about observational representativeness. Adding to these concerns, scales were not clearly reported in most studies, suggesting that it is a minor consideration. Journals can help mitigate this problem by implementing scale reporting standards, which can spur ecologists to more rapidly adopt new observational technologies, and thereby close key gaps in current observational domains.}

@@

Our analysis focused on two dimensions of spatial scale, resolution (grain) and extent, and two of temporal scale, interval and duration (Table 1, and see SI for full definitions). Resolution is the area of an individual spatial replicate within which a complete measurement (as opposed to a sub-sample) of the feature of interest was made. Extent is the area enclosed by the outer-most spatial replicates, or, if the system or habitat being sampled was distinct from its surrounding matrix (e.g. forest patches in grassland habitats), the summed area of sampled patches. Interval refers to the average time elapsed between individual temporal replicates. Duration measures the time elapsed between the first and last temporal replicates, or, in the case of temporally unreplicated observations, the estimated time spent collecting the observation. We also assessed observational scales within two additional dimensions, \emph{actual} extent (the integrated area of spatial replicates) and \emph{actual} duration (the summed observational time of temporal replicates). We evaluated these additional dimensions to gain insight into how much the actual scales of observation (i.e. how much space and time is covered by the measurement) differ from those they are intended to represent.

@@

\noindent \textbf{Results}\\
\noindent \textbf{\emph{Domains}}\\
In terms of resolution, the majority (67\%) of observations were collected in plots of $<$1 m$^2$ resolution, 24\% were collected within plots of 1 m$^2$ up to 1 ha, and the remaining 9\% in plots of $\geq$1 ha (Fig. 1A). The extent of 19\% of observations was $<$10 ha, 23\% covered 10-1,000 ha, 11\% 1,000-10,000 ha, 19\% 10,000-100,000 ha, 12\% 100,000-1,000,000 ha, and 16\% $>$1,000,000 ha (Fig. 1B).

@@

Comparing the two spatial dimensions against one another (for all observations) shows a primary concentration of observations with 10 cm$^2$ to 100 m$^2$ resolution that have extents ranging between slightly over 1,000 to nearly 1,000,000 ha (Fig. 2C). The second-most prominent concentration consists of higher resolution (1 cm$^2$-1 m$^2$), smaller extent (10-1,000 ha) observations, beneath which lies a third and fainter concentration of 1-1,000 cm$^2$ resolution, 1000 m$^2$ to $<$10 ha.  These three concentrations suggest a tendency for observational extent to increase with resolution, which is a relationship that becomes more pronounced in the less densely observed portion of the domain where resolutions $\geq$100 m$^2$.

Most temporally replicated observations have daily to decadal intervals and durations of $\geq$ 1 month to 1 decade. The orientation of this concentration shows that interval increases with duration; observations lasting one month to one year tend to have daily to monthly intervals, while those lasting one year to one decade tend to have yearly to decadal intervals. The low densities of observations having sub-daily intervals shows that relatively few high frequency, long duration ecological measurements are undertaken.

To provide further insight into observational domains, we also evaluated the degree to which the scales \emph{represented} by observations differed from those that were actually observed. To make this assessment, we first log$_{10}$ transformed and then subtracted the values of i) actual extent from extent and ii) actual duration from duration, in order to calculate the magnitude of difference (or decade) between each pair of dimensions for each observation. We then compared how these magnitudes varied in relation to the scale of the actual dimension (Fig. 3). This comparison showed that most (81\%) of the assessed observations had actual extents of $\leq$1 ha, which on average were 4 to nearly 8 orders of magnitude smaller than the quantified extent (Fig 3.A).  Actual extent converged with extent $\geq$100,000 ha, but this applies to just 5\% of observations. The actual duration of 64\% of observations was $<$ 1 day, which on average was 3 to nearly 9 orders of magnitude shorter than the time span covered by temporal replicates (Fig. 3B). The two duration measures were approximately equal for the 16\% of observations exceeding one month of actual duration.

@@

\noindent \textbf{\emph{Potential biases and uncertainties in quantifying scales}}\\
There were several potential methodological aspects that could have influenced our assessment of ecology's spatial and temporal domains. The first stems from our finding that many studies did not precisely report observational scales, which meant that we had to estimate, rather than simply record, these values for most observations (specifically, in 63\%, 60\%, and 69\% of cases for resolution, extent, and actual extent, and 36\%, 64\%, and 83\% of cases for interval, duration, and effective duration, respectively). The inevitable estimation errors may have biased our overall findings. However, we attempted to quantify and account for this error by assessing inter-rater disagreement and incorporating this uncertainty into our resampling methodology. The resulting confidence intervals (Fig. 1) suggest that it was unlikely that estimation errors unduly influenced our findings.

@@

%Adding to our motivation to record actual observational scales was the fact that 
\noindent \textbf{Discussion}\\
\noindent \textbf{\emph{Insights into the scale domains of modern ecology}}\\
Our results suggest that the scale domains of modern ecological observations are fairly narrow, and are collected primarily with conventional field techniques. Spatially, the majority of observations have grains of $\leq$1 m$^2$ and extents of $\leq$10,000 ha (Fig 1A;B). In the temporal domains, most observations are either un-replicated snapshots, or of low frequency ($\geq$1 month interval; Fig. 1C) and relatively short duration ($\leq$1 year; Fig. 1D; 2D). Contrasting observational dimensions reveals that larger extents are associated with larger plot sizes (Fig. 2C), while longer durations are associated with longer intervals (Fig. 2D). The latter association presumably reflects a cost-imposed tradeoff between sampling frequency and temporal duration that is characteristic of traditional field-based observation. The same tradeoff is also responsible for the inverse relationship between resolution and interval that dominates that domain space (Fig. 2A). As a result of these tradeoffs, there are notable ``holes'' in the domains defined by high frequency (daily to sub-daily intervals) observations having 1) high to moderate resolutions ($\geq$1 m$^2$ up to 100 ha; Fig. 2A) and 2) decadal or longer durations (Fig. 2B).

@@

Have these domains changed since the seminal papers on scale first began to appear in the late 1980s \cite{wiens_spatial_1989, levin_problem_1992, tilman_ecological_1989}? A comprehensive answer to this question would require a similar study focused on earlier literature, but an analysis of results presented in three prior studies provides partial insight. The first and most comprehensive dataset consists of duration values extracted by Tilman \cite{tilman_ecological_1989} from 623 studies published between 1977-1987 in the journal \emph{Ecology}. The average duration of the most comparable subset of those values (n=419; see SI) was 3.6 years, compared to 3.3 years for observations in our sample (or 5.1 if temporally un-replicated observations are excluded). The second dataset is found in Kareiva and Anderson \cite{kareiva_spatial_1988}, who present the resolutions of 97 community ecology experiments published in \emph{Ecology} between 1980-1986. The average of those (12,657 m$^2$) was substantially smaller than the mean of our sample (1,510,247 m$^2$), but comparing the 80th percentile value (197 m$^2$) of Kareiva and Anderson's \cite{kareiva_spatial_1988} to ours (115 m$^2$) shows that the majority of contemporary observations are finer-grained than most 1980s-era experiments. The third dataset is provided by Porter et al \cite{porter_wireless_2005}, who compared the extent and interval of 25 studies published in 2003 and 2004 (also in \emph{Ecology}). The mean interval was 178 days, compared to 684 days in our sample, but the 80th percentile value in our study was 169 days compared to 329 days in theirs. Extent in our sample was substantially larger according to multiple summary statistics, including the mean (115,098,283 ha in our study versus 368,403 ha), median (5,120 ha versus 9 ha), and 95th percentile (45,104,594 ha versus 136,000 ha).

Although limited due to methodological differences (e.g. a focus on experiments versus un-manipulated systems), these comparisons suggest that the duration and, less clearly, resolution of ecological observations have changed little in the past 30 years, but observational frequency and particularly extent have increased. The growth in observational extent is also reflected by the positive year-on-year trend within our own dataset, which corresponds to the increasing use of remote sensing (see figures in SI).

@@

It is therefore important to examine whether the phenomena being observed are adequately captured by the design of replicates. Our methods suggest one possible procedure for assessing the representativeness of replicates: 1) measure auto-correlation (spatial or serial) in the replicates, 2) add the autocorrelation length to the replicate area/duration, 3) calculate an autocorrelation-adjusted actual extent/duration, and 4) plot where it falls between actual extent/duration and extent/duration. If the adjusted actual value is substantially closer to the representative value, then this may provide greater confidence in the adequacy of the sampling scheme. If not, then alternative sampling methods may be used to close this gap. For example, remote sensing provides wall-to-wall spatial coverage of a study area, erasing the difference between actual extent and extent. Furthermore, the interval of high-resolution imaging (higher resolution is preferred in images as it allows individual features to be better discerned \cite{dark_modifiable_2007, hay_comparison_2003}) is now approaching daily to sub-daily scales \cite{drusch_sentinel-2:_2012, hand_startup_2015}, allowing improved representation of spatial and temporal dynamics. For phenomena that can't be measured from space, either because they are not visible or because they require continuous observation, new approaches for collecting \emph{in situ} or near-surface observations (e.g. low-cost wireless sensors \cite{wolf_gsm-based_2012, collins_new_2006, porter_wireless_2005}, citizen observers \cite{dickinson_current_2012}, and autonomous vehicles \cite{anderson_lightweight_2013}) can be used to increase the spatial and temporal coverage of observations.
